Measuring the double-drift illusion with hand tracking

### 2021-11-19 Manuscript.Rmd
work on revisions

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived position deviates further and further away from  its true path. We first evaluated the illusion using manual tracking. Participants followed the gabor with a stylus on a drawing tablet that coincided optically with the horizontal monitor surface. Their hand and the stylus were not visible during the tracking. The magnitude of the tracking illusion corresponded closely to previous perceptual and pointing measures indicating that manual tracking is a valid measure for the illusion. This allowed us to use it in a second experiment to capture the behavior of the illusion as it eventually degrades and breaks down in single trials. Specifically, the deviation of the gabor stops accumulating at some point and either stays at a fixed offset or resets toward the veridical position. To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor. Resets were detected and analyzed but their distribution neither matches a strict temporal or spatial limit on perception that could trigger resets.

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves, as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived location continues to move away from its physical position by up to several times the size of the gabor. The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021]. Rather than use point estimates of the perceived location of the stimulus, we want to test how well a continuous trajectory can measure the strength of the illusion, or if perception is changed by this response modality such as in other illusions

CITE SOME PREVIOUS WOKR HERE.

We assess this in our first experiment, using hand movement trajectores recorded during stimulus presentation (Fig. 1). We compare trajectories across 5 different internal motions of the gabor and compare the illusion strength with previous work based on a straightforward vector combination model [@Heller2021]. As we expected [@mhjh], online tracking provides a similar measure of illusion strength as pointing.

In our second experiment, we look at the spatial and temporal limits of the illusion. At some point, the illusion stops accumulating. The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far. Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path. Others report that the position moves back toward its physical location, either slowly or suddenly, whereupon accumulation begins again. The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.

@@

@@

```{r fig3, echo=F, fig.height=3.4125, fig.width=6, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets. **Left:** At some point, the perceived position may stop moving further away from its actual path. For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two. **Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), either due to stimulus properties or within or between subject differences, resets may occur at different points in space. **Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path. **Bottom right:** if they occur instead after some period of time, resets should have a constant radius (R); the time since the gabor’s motion began."}
library(magick)
Fig3 <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig3)
```

@@

For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were in accordance with the Declaration of Helsinki (2013) and were approved by York’s Human Participants Review Committee. All participants provided prior, written, informed consent.

@@

In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus (data not used, but corresponds to illusion strength determined from trajectories). More details are given below. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. The re-tracing task was done in half of 8 blocks (the other half of the blocks are not used). Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block. Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials in total.

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes. First, we remove the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that go toward the participant, to remove any unintended jitters caused by starting and ending the movement. Then, we apply a bi-directional low-pass Butterworth filter with a cut-off frequency of 1.5 Hz. However, since participants draw at very different speeds, we first linearly interpolate the trajectories such that the sampling rate is normalized to 30 Hz for all trajectories, based on the stimulus duration. In these summary trajectories we detect the first inflection point using the X coordinates. This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate, and finally locate the sample in the raw trajectory that is closest to this point. Not all trajectories result in a reset point, so we may miss some genuine hit-the-wall resets where the trajectory does not have an inflection point. However, there is no simple way to segregate these from trials without a reset.

>>>>

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived position deviates further and further away from  its true path. We first evaluated the illusion using manual tracking. Participants followed the gabor with a stylus on a drawing tablet that coincided optically with the horizontal monitor surface. Their hand and the stylus were not visible during the tracking. The magnitude of the tracking illusion corresponded closely to previous perceptual and pointing measures indicating that manual tracking is a valid measure for the illusion. This allowed us to use it in a second experiment to capture the behavior of the illusion as it eventually degrades and breaks down in single trials. Specifically, the deviation of the gabor stops accumulating at some point and either stays at a fixed offset or resets toward the veridical position. To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor. Resets were detected and analyzed but their distribution neither matches a temporal nor a spatial limit on perception that could trigger resets.

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves, as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived location continues to move away from its physical position by up to several times the size of the gabor.

The strength of the illusion; the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021]. Rather than use point estimates of the perceived location of the stimulus, we want to test how well a continuous trajectory can measure the strength of the illusion. Eye and hand movement are differentially sensitive to many illusions, such as the Ponzo illusion [@Gamble2017], Ebbinghaus illusion [@Knol2017], flash-lag effect [@Blohm2003], the Duncker illusion [@Soechting2001] and many others. Since it has been shown that the double-drift illusion affects pointing movements, but not saccades [@Lisi2017] we expect that hand trajectories will also be sensitive to the illusion.

We assess this in our first experiment, using hand movement trajectores recorded during stimulus presentation (Fig. 1). We compare trajectories across 5 different internal motions of the gabor and compare the illusion strength with previous work based on a straightforward vector combination model [@Heller2021]. As we expected, online tracking provides a similar measure of illusion strength as pointing.

In our second experiment, we look at the spatial and temporal limits of the illusion. At some point, the illusion stops accumulating. The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far. Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path. Others report that the position moves back toward its physical location, either slowly or suddenly, whereupon accumulation begins again. The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.

@@

In particular, we compare the illusion strengths we find with those reported by Heller et al. [-@Heller2021]. They show a double-drift stimulus with 36 combinations of internal and external motion for 500 ms and then have participants indicate on a ring where they perceived the path of the stimulus to intersect with this ring. While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will mostly use their overall estimate of illusion strength, given by $K=0.74$.

@@

```{r fig3, echo=F, fig.height=3.4125, fig.width=6, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets. **Left:** At some point, the perceived position may stop moving further away from its actual path. For some people this takes the form of a ‘return’ reset, back toward the true position either suddenly or slowly, for others it means the illusory position remains at a fixed offset moving parallel to the real path. **Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), e.g. due to stimulus properties or within or between subject differences, resets may occur at different points in space. **Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path. **Bottom right:** if they occur instead after some period of time, resets should have a constant radius (R); the time since the gabor’s motion began."}
library(magick)
Fig3 <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig3)
```

@@

For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee. All participants provided prior, written, informed consent.

@@

In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus (data not used, but corresponds to illusion strength determined from trajectories). More details are given below. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. The re-tracing task was done in half of 8 blocks (the other half of the blocks are not used). Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block. Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials for each participant, and 1296 trials in total.

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes. First, we remove the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that go toward the participant, to remove any unintended jitters caused by starting and ending the movement. Then, we apply a bi-directional low-pass Butterworth filter with a cut-off frequency of 1.5 Hz. However, since participants draw at very different speeds, we first linearly interpolate the trajectories such that the sampling rate is normalized to 30 Hz for all trajectories, based on the stimulus duration. In these summary trajectories we detect the first inflection point using the X coordinates. This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate, and finally locate the sample in the raw trajectory that is closest to this point. Finally, we remove a few reset points that are detected very close to the veridical path without any build-up of the illusion by only keeping reset points that fell 5 mm within the quadrant where we expect them. This heuristic will result in some misses and some false alarms, and while changing the algorithm may improve performance, there is no ground truth to evaluate this. All raw trajectories and the detected reset points can be found on the OSF repository (XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX add link?).
